Well-quasi-orderings on word languages

### 2025-10-13 src/introduction.tex
m

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}, even in the simplified
setting of a finite alphabet $\Sigma$ equipped with the equality relation. In
this setting, the three alternatives we consider are the \intro{prefix
relation} ($u \intro*\prefleq v$ if there exists $w$ with $uw = v$), the
\intro{suffix relation} ($u \intro*\suffleq v$ if there exists $w$ such that
$wu = v$), and the \intro{infix relation} ($u \intro*\infleq v$ if there exists
$w_1,w_2$ such that $w_1 u w_2 = v$). Note that these three relations
straightforwardly generalize to infinite quasi-ordered alphabets.
Unfortunately, it is easy to see that none of these constructions are 
well-quasi-ordered as soon as the alphabet contains two distinct letters:
for instance, the infinite sequence $a^n b^n a^n$ is \kl{well-quasi-ordered} by
the \kl{subword relation} but by neither the \kl{prefix relation}, nor the
\kl{suffix relation}, nor the \kl{infix relation}.

@@

A similar question was studied by Atminas, Lozin, and Moshkov \cite{ALM17}, in
the hope of finding characterizations of classes of \emph{finite graphs} that
are \kl{well-quasi-ordered} by the \emph{induced subgraph relation}
\cite[Section 7]{ALM17}. In this setting, it is common to refer to classes of
graphs via a list of \emph{forbidden patterns}, which are finite graphs that
cannot be found as induced subgraphs in the class. Applying this reasoning to
finite words with the \kl{infix relation}, they provide an efficient decision
procedure for checking whether a language $L \subseteq \Sigma^*$ is
\kl{well-quasi-ordered} by the \kl{infix relation} whenever said language is
given as input via a list of \emph{forbidden factors} \cite[Theorem 1, Theorem
2]{ALM17}. The key construction of their paper is to study languages $L$ that
are \emph{regular} (recognized by some finite deterministic automata), for
which they can decide whether $L$ is \kl{well-quasi-ordered} by the \kl{infix
relation} \cite[Theorem 1]{ALM17}. Because it is easy to transform a list of
forbidden factors into a regular language \cite[Theorem 1]{ALM17}, this yields
the desired decision procedure. Our work extends this result in several ways:
first, we also consider the \kl{prefix relation} and the \kl{suffix relation},
then we consider non-regular languages, and finally, we provide very precise
descriptions of the \kl{well-quasi-ordered} languages, as well as tight bounds
on their \kl{ordinal invariants}. In particular, we highlight the role of
\kl{bounded languages} as the one that are well-behaved with respect to being
\kl{well-quasi-ordering} by the \kl{infix relation} \cref{infixes-bounded:sec}.

>>>>

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}, even in the simplified
setting of a finite alphabet $\Sigma$ equipped with the equality relation. In
this setting, the three alternatives we consider are the \intro{prefix
relation} ($u \intro*\prefleq v$ if there exists $w$ with $uw = v$), the
\intro{suffix relation} ($u \intro*\suffleq v$ if there exists $w$ such that
$wu = v$), and the \intro{infix relation} ($u \intro*\infleq v$ if there exists
$w_1,w_2$ such that $w_1 u w_2 = v$). Note that these three relations
straightforwardly generalize to infinite quasi-ordered alphabets.
Unfortunately, it is easy to see that none of these constructions are 
well-quasi-ordered as soon as the alphabet contains two distinct letters:
for instance, the infinite sequence $a b^n a$ is \kl{well-quasi-ordered} by
the \kl{subword relation} but by neither the \kl{prefix relation}, nor the
\kl{suffix relation}, nor the \kl{infix relation}.

@@

A similar question was studied by Atminas, Lozin, and Moshkov \cite{ALM17}, in
the hope of finding characterizations of classes of \emph{finite graphs} that
are \kl{well-quasi-ordered} by the \emph{induced subgraph relation}
\cite[Section 7]{ALM17}. In this setting, it is common to refer to classes of
graphs via a list of \emph{forbidden patterns}, which are finite graphs that
cannot be found as induced subgraphs in the class. Applying this reasoning to
finite words with the \kl{infix relation}, they provide an efficient decision
procedure for checking whether a language $L \subseteq \Sigma^*$ is
\kl{well-quasi-ordered} by the \kl{infix relation} whenever said language is
given as input via a list of \emph{forbidden factors} \cite[Theorem 1, Theorem
2]{ALM17}. The key construction of their paper is to study languages $L$ that
are \emph{regular} (recognized by some finite deterministic automata), for
which they can decide whether $L$ is \kl{well-quasi-ordered} by the \kl{infix
relation} \cite[Theorem 1]{ALM17}. Because it is easy to transform a list of
forbidden factors into a regular language \cite[Theorem 1]{ALM17}, this yields
the desired decision procedure. Our work extends this result in several ways:
first, we also consider the \kl{prefix relation} and the \kl{suffix relation},
then we consider non-regular languages, and finally, we provide very precise
descriptions of the \kl{well-quasi-ordered} languages, as well as tight bounds
on their \kl{ordinal invariants}. In particular, we highlight the central role of
\kl{bounded languages} as the one that are well-behaved with respect to being
\kl{well-quasi-ordering} by the \kl{infix relation} \cref{infixes-bounded:sec}.\todo[inline]{
 i tried to rewrite it but im not sure i understand this sentence :)
}
